Text embedding models enable high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-30 paper/main.tex
backing up progress, rewrote framing of question-based analyses/results

%\title{Text embedding models reveal high-resolution insights into conceptual
%knowledge and learning from short multiple-choice quizzes}
\title{A geometric framework for capturing high-resolution insights into
conceptual knowledge and learning in classroom-like settings}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge in real-world educational contexts. Our approach embeds
course content in a high-dimensional
conceptual space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
watched two lecture videos from the Khan Academy platform, interleaved
between three short multiple-choice quizzes. We applied our
framework to the videos' transcripts and the text of the quiz questions to
quantify the conceptual content presented in each moment of video and knowledge probed by each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings demonstrate how a
small set of quiz questions may be used to obtain rich and meaningful
high-resolution insights into individual students' knowledge, and how it
changes over time as they learn.

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model resulted in a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics learned by the model). Within this space, each lecture's series of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}A). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

Next, wondered whether we could use this same model to capture the

>>>>

\title{Text embedding models enable high-resolution insights into
conceptual knowledge and learning in classroom-like settings}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge in real-world educational contexts. Our approach embeds
course content in a high-dimensional
conceptual space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
viewed two lecture videos from the Khan Academy platform, interleaved
between three short multiple-choice quizzes. We applied our
framework to the lectures' transcripts and the text of the quiz questions to
quantify the conceptual content presented in each moment of video and knowledge probed by each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings demonstrate how a
small set of quiz questions may be used to obtain rich and meaningful
high-resolution insights into individuals' knowledge, and how it
changes over time as they learn.

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model yielded a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics discovered by the model). Within this space, each lecture's sequence of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

Next, we sought to characterize what information participants' performance on each of the three quizzes could provide about their conceptual knowledge at that point in time. Traditional approaches to evaluating students' performance on short-form knowledge assessments such as those in our study entail computing the proportion of correctly answered questions to assign a simple numeric score. While this may provide some measure of the \textit{extent} of a learner's knowledge, such a score affords little information to either the teacher or the learner about the particular \textit{contents} of their knowledge---in other words, raw ``proportion-correct'' measures may capture \textit{how much} a student knows, but not \textit{what} they know. For instance, suppose a participant in our study was highly knowledgeable about fundamental physical forces (i.e., lecture 1 content) but unfamiliar with how stars are formed (i.e., lecture 2 content), while a second participant was unfamiliar with fundamental forces but had extensive prior knowledge about star formation. Since quiz 1 (completed before viewing either lecture) contained an equal number of questions about each lecture's content (see Fig.~\ref{fig:experiment}), these two participants could easily achieve the same proportion-correct score despite their conceptual knowledge differing substantially. How might we distinguish these individuals?

We hypothesized that our text embedding model, which we had trained on transcripts of the two lectures to characterize their conceptual contents, should also allow us to capture the conceptual knowledge probed by each quiz question. Given that the questions were (ostensibly, by design) ``about'' the content of

In other words, given that the questions were (ostensibly, by design) ``about'' one lecture or the other,
